---
title: "DoIt for a bimodal 2-d density"
author: "Stefan Siegert"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{doit_bimodal_example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r setup, include = FALSE}
knitr::opts_chunk$set(
  fig.path = paste('figure/', knitr::current_input(), '-figs/', sep=''),
  fig.width = 7,
  fig.height = 5
)
```


```{r, eval=FALSE, echo=TRUE}
library(doit)
library(tidyverse)
library(mvtnorm)
library(viridis)
library(lhs)
```
```{r, eval=TRUE, echo=FALSE}
suppressPackageStartupMessages(library(doit))
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(mvtnorm))
suppressPackageStartupMessages(library(viridis))
suppressPackageStartupMessages(library(lhs))
```



## The target function


We use the DoIt method to approximate a 2-dimensional bimodal
density which is proportional to a weighted sum of 2 bivariate
Normal distributions:

```{r}
f_target = function(x) {
  2/3 * dmvnorm(x, c(0,0), diag(c(1,1))) + 
  1/3 * dmvnorm(x, c(3,3), matrix(c(2,.6, .6, 1),2,2))
}

df_true = crossing(x1 = seq(-5, 10, .1), x2 = seq(-5,10,.1)) %>%
  mutate(f = f_target(cbind(x1,x2)))

head(df_true)
```

```{r plot-true}
ggplot(df_true) + 
  geom_raster(aes(x=x1, y=x2, fill=f)) + 
  geom_contour(aes(x=x1, y=x2, z=f), col='black', lty=2) + 
  scale_fill_viridis()
```


# Latin hypercube design

We sample a latin hypercube design on the interval `[-5,10]` in both dimensions:

```{r}
set.seed(123)
design = maximinLHS(n=100, k=2) %>% 
         as_data_frame %>% 
         setNames(c('x1', 'x2')) %>%
         mutate(x1 = x1 * 15 - 5, 
                x2 = x2 * 15 - 5,
                f  = f_target(cbind(x1, x2)))
head(design)
```

```{r plot-design}
ggplot(design) + 
  geom_point(aes(x=x1, y=x2, colour=f), cex=3) + 
  scale_colour_gradient2()
```



# Fit DoIt approximation

We use `doit_estimate_w()` to estimate the optimal kernel width for
our design. We then use `doit_fit()` to estimate the parameters for
the approximation, and `doit_approx()` to approximate the target
function at the desired parameter values:

```{r}
w          = doit_estimate_w(design)
doit       = doit_fit(design, w=w)
x_eval     = df_true %>% select(x1, x2)
df_approx  = doit_approx(doit, x_eval)
```

We plot the approximation and compare it to the target function: 

```{r plot-approx}
plot_df = bind_rows(df_approx %>% rename(f=dens_approx) %>% 
                                  mutate(type='approx'),
                    df_true %>% mutate(type='true'))
ggplot(plot_df, aes(x=x1, y=x2)) + 
  geom_raster(aes(fill=f)) + 
  facet_wrap(~type) +
  geom_point(data=design, pch=1, colour='white') + 
  scale_colour_viridis() + scale_fill_viridis()
```

Even though only very few design points are in the "active region"
of the target function, the approximation looks quite accurate.

We also look at the variance of the approximation:

```{r plot-variance}
ggplot(df_approx, aes(x=x1, y=x2)) + 
  geom_raster(aes(fill=variance)) + 
  geom_contour(aes(z=dens_approx), col='black', lty=2) + 
  geom_point(data=design, pch=1, colour='white') + 
  scale_colour_viridis() + scale_fill_viridis()
```

The approximation variance is highest in between design points
inside the active region. These high variance regions will be
targeted when adding new design points during sequential updating.


## Sequential update of the design

Here we sequentially add 25 new design points, reestimating the
kernel width every 5th iteration:

```{r}
n_new = 25
for (jj in 1:n_new) {
  x_n        = doit_propose_new(doit)
  design_add = x_n %>% mutate(f = f_target(x_n))
  design     = bind_rows(design, design_add)
  if (jj %% 5 != 0) {
    doit = doit_update(doit, design_add)
  } else { 
    w    = doit_estimate_w(design, w_0=doit$w)
    doit = doit_fit(design, w=w)
  }
}
```

We plot the updated approximation and highlight the new design
points:

```{r plot-design-update-25}
df_approx = doit_approx(doit, x_eval)
ggplot(mapping=aes(x=x1, y=x2)) +
  geom_raster(data=df_approx, aes(fill=dens_approx)) + 
  geom_point(data=design %>% 
                  mutate(design=rep(c('latin_hypercube','sequential'), c(n()-n_new, n_new))),
             mapping=aes(pch=design, colour=design), cex=2) + 
  scale_fill_viridis()
```

Lastly, we use `doit_marginals()` to approximate the marginal densities:

```{r plot-marg}
df_marg = doit_marginals(doit, x_eval)
ggplot(df_marg, aes(x=theta, y=dens_approx)) + 
  geom_point() + geom_line() + facet_wrap(~par)
```



