---
title: "DoIt for a 2-d example"
author: "Stefan Siegert"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{doit_2d_example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r setup, include = FALSE}
knitr::opts_chunk$set(
  fig.path = paste('figure/', knitr::current_input(), '-figs/', sep=''),
  fig.width = 7,
  fig.height = 5
)
```


```{r, eval=FALSE, echo=TRUE}
library(doit)
library(tidyverse)
library(mvtnorm)
library(viridis)
library(lhs)
```
```{r, eval=TRUE, echo=FALSE}
suppressPackageStartupMessages(library(doit))
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(mvtnorm))
suppressPackageStartupMessages(library(viridis))
suppressPackageStartupMessages(library(lhs))
```


## The target function


We use the DoIt method to approximate the banana-shaped density used in Joseph
(2012):

```{r}
f = function(theta) {
  theta = as.matrix(theta)
  theta = matrix(theta, ncol=2) 
  x = cbind(theta[ , 1, drop=FALSE], 
            theta[ , 2, drop=FALSE] + 0.03 * theta[ , 1, drop=FALSE] ^ 2 - 3)
  dmvnorm(x, c(0,0), diag(c(100, 1)))
}

df_true = crossing(theta1 = seq(-30, 30, .5), theta2 = seq(-12,7,.5)) %>%
  mutate(f = f(cbind(theta1,theta2)))

head(df_true)
```

```{r plot-true}
ggplot(df_true) + 
  geom_raster(aes(x=theta1, y=theta2, fill=f)) + 
  geom_contour(aes(x=theta1, y=theta2, z=f), col='black', lty=2) + 
  scale_fill_viridis()
```


## Latin hypercube design

We evaluate the function at 100 parameter values on a maximin latin hypercube
design. As the approximation region we choose the interval `[-27.5, 27.5]` in
the `theta1` direction, and `[-10,5]` in the `theta2` direction.

```{r}
lim1 = data_frame(min=-27.5, max=27.5, range=max-min)
lim2 = data_frame(min=-10,   max=5,    range=max-min)
set.seed(123)
design = maximinLHS(n=100, k=2) %>% 
         as_data_frame %>% 
         setNames(c('theta1', 'theta2')) %>%
         mutate(theta1 = theta1 * lim1$range + lim1$min, 
                theta2 = theta2 * lim2$range + lim2$min,
                f      = f(cbind(theta1, theta2)))
head(design)
```

```{r plot-design}
ggplot(design) + 
  geom_point(aes(x=theta1, y=theta2, colour=f), cex=3) + 
  scale_colour_gradient2()
```


## Estimate the optimal kernel width

The DoIt method works by approximating the target density by a weighted sum of
Normal distributions. The optimal variance parameter `w` of the Normal
distributions is estimated by the function `doit_estimate_w()`:


```{r}
w = doit_estimate_w(design)
print(w)
```


## Evaluate the DoIt approximation

The function `doit_fit()` estimates the DoIt parameters and saves them in an
object of class `doit`:

```{r}
doit = doit_fit(design, w=w)
str(doit)
```


We specify the parameter values at which we want to approximate the target
density, and use `doit_approx()` to calculate the approximation:

```{r}
theta_eval = df_true %>% select(theta1, theta2)
df_approx = doit_approx(doit, theta_eval)
```


```{r plot-approx}
plot_df = bind_rows(df_approx %>% rename(f=dens_approx) %>% 
                                  mutate(type='approx'),
                    df_true %>% mutate(type='true'))
ggplot(plot_df, aes(x=theta1, y=theta2)) + 
  geom_raster(aes(fill=f)) + 
  facet_wrap(~type) +
  geom_point(data=design, pch=1, colour='white') + 
  scale_colour_viridis() + scale_fill_viridis()
```


The function `doit_approx()` also returns the estimation error variance:

```{r plot-variance}
ggplot(df_approx, aes(x=theta1, y=theta2)) + 
  geom_raster(aes(fill=variance)) + 
  geom_point(data=design, pch=1, colour='white') + 
  scale_colour_viridis() + scale_fill_viridis()
```


## Add more design points

To add further points to the existing design, we use the function
`doit_propose_new()` which chooses a new design point where the current
estimation variance is particularly high:

```{r}
theta_n = doit_propose_new(doit)
print(theta_n)
```

```{r plot-design-update-1}
ggplot(df_approx, aes(x=theta1, y=theta2)) + 
  geom_raster(aes(fill=variance)) + 
  scale_fill_viridis() +
  geom_point(data=design, pch=1, col='white') +
  geom_point(data=theta_n, pch=16, cex=3, col='red')
```



We could use `doit_fit()` to estimate the DoIt parameters again on the original
design plus the added point. For more efficiency, we can use the function
`doit_update()`. Note that the kernel width `w` is not updated. In the
following we add 75 design points, re-optimising the kernel width `w` every
10th iteration: 

```{r}
n_new = 75
for (jj in 1:n_new) {
  theta_n    = doit_propose_new(doit)
  design_add = theta_n %>% mutate(f = f(theta_n))
  design = bind_rows(design, design_add)
  if (jj %% 10 != 0) {
    doit = doit_update(doit, design_add)
  } else { 
    w = doit_estimate_w(design, w_0=doit$w)
    doit = doit_fit(design, w=w)
  }
}
```


Here we plot the new approximation and highlight the sequentially added new
design points:

```{r plot-design-update-75}
theta_eval = df_true %>% select(theta1, theta2)
df_approx = doit_approx(doit, theta_eval)
ggplot(mapping=aes(x=theta1, y=theta2)) +
  geom_raster(data=df_approx, aes(fill=dens_approx)) + 
  geom_point(data=design %>% 
                  mutate(design=rep(c('latin_hypercube','sequential'), c(n()-n_new, n_new))),
             mapping=aes(pch=design, colour=design), cex=2) + 
  scale_fill_viridis()
```


## Marginal densities

The function `doit_marginals()` returns a data frame containing approximations
of all marginal distributions:

```{r}
df_marg = doit_marginals(doit, theta_eval)
ggplot(df_marg, aes(x=theta, y=dens_approx)) + 
  geom_point() + geom_line() + 
  facet_wrap(~par, scales='free')
```

To calculate only the approximate marginal of, say, `theta1`, use
`doit_marginal()`. If the argument `theta_eval` is unspecified, the design
values are used:

```{r}
df_marg_theta1 = doit_marginal(doit, 'theta1')
head(df_marg_theta1)
```


## Expectations and variances

We use the functions `doit_expectation()` and `doit_variance()` to approximate
the expectation and covariance matrix of the target density:

```{r}
doit_expectation(doit)
```

```{r}
doit_variance(doit)
```


