---
title: "DoIt for a 1-d example"
author: "Stefan Siegert"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{doit-1d-example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r setup, include = FALSE}
knitr::opts_chunk$set(
  fig.path = paste('figure/', knitr::current_input(), '-figs/', sep=''),
  fig.width = 7,
  fig.height = 5
)
```


```{r, echo=TRUE, eval=FALSE}
library(doit)
library(lhs)
library(tidyverse)
```
```{r, echo=FALSE, eval=TRUE}
suppressPackageStartupMessages(library(doit))
suppressPackageStartupMessages(library(lhs))
suppressPackageStartupMessages(library(tidyverse))
```

## The target distribution

We use the example model from Joseph (2012), namely a Bernoulli likelihood with
logit link function and a Normal prior: 

- $r \sim N(\mu, \tau^2)$
- $p = \frac{1}{1+exp(-r)}$
- $y \sim Ber(p)$
- where $y=1$, $\mu = 1$, $\tau = 4$

The unnormalised posterior is given by

```{r}
h = function(r) {
  p = 1/(1+exp(-r))
  dbinom(1, 1, p) * dnorm(r, 1, 4)
}
```


## Design points

We evaluate the unnormalised posterior at 10 equally spaced design points:

```{r}
design = data.frame(x=seq(-10, 20, len=10))
design$f = sapply(design$x, h)
```


## Estimate the optimal kernel width


The DoIt method works by approximating the target density by a weighted sum of
Normal distributions. The optimal variance parameter `w` of the Normal
distributions is estimated by the function `doit_estimate_w()`:

```{r}
w = doit_estimate_w(design)
print(w)
```


## Evaluate the DoIt approximation

The function `doit_fit()` estimates the parameters of the DoIt method and saves
them in an object of class `doit`:

```{r}
doit = doit_fit(design, w=w)
```

We specify the parameter values at which we want to approximate the target
posterior and use `doit_approx()` to calculate the approximation:

```{r}
theta_eval = data.frame(x = seq(-15,25,.1))
doit_out = doit_approx(doit, theta_eval)
```

The approximation function returns the approximated density value and the
estimation error variance:

```{r}
head(doit_out)
```

The output of `doit_approx()` works nicely with the `ggplot2` package:

```{r plot-doit-approx}
ggplot() + 
geom_line(data=doit_out, aes(x=x, y=dens_approx, colour=variance)) +
geom_point(data=design, aes(x=x, y=0), pch=1, cex=2)
```


## Add more design points

To extend the existing design, we use `doit_propose_new()` to choose a new
design point where the current estimation variance is particularly high:

```{r}
x_n = doit_propose_new(doit)
design_new = mutate(x_n, f=h(x))
```

Instead of using `doit_fit()` again on the original design plus additional new
point, we can use the `doit_update()` function which is more efficient. Note
that the kernel width `w` is not updated:

```{r}
doit = doit_update(doit, design_new)
```

We plot the new approximation and highlight the added design point:

```{r plot-doit-approx-update}
doit_out = doit_approx(doit, theta_eval)
ggplot() + 
geom_point(data=design_new, aes(x=x, y=0), pch=1, cex=2, col='red') +
geom_point(data=design, aes(x=x, y=0), pch=1, cex=2) +
geom_line(data=doit_out, aes(x=x, y=dens_approx, colour=variance))
```


## Posterior expectation and variance

We have functions to approximate posterior mean and variance:

```{r}
doit_expectation(doit)
doit_variance(doit)
```


